---
output: github_document
---


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# treenomial

[![CRAN status](https://www.r-pkg.org/badges/version/treenomial)](https://cran.r-project.org/package=treenomial)

## Overview
 
The package **treenomial** is an application of polynomials that uniquely describe trees. It provides tools for tree analysis and comparison based on polynomials. The core functions are: 

* __`treeToPoly()`__: convert rooted unlabeled binary trees to tree distinguishing polynomials described with coefficient matrices

* __`polyToDistMat()`__: construct a distance matrix from multiple coefficient matrices using a distance measure 

For the mathematical description of the tree defining polynomial see: 

[Liu, Pengyu. "A tree distinguishing polynomial." arXiv preprint arXiv:1904.03332 (2019).](https://arxiv.org/abs/1904.03332)


## Installation

To install using [CRAN](https://cran.r-project.org/package=treenomial): 

```
install.packages("treenomial")
```

For the development version:

```
library(devtools)
install_github("mattgou1d/treenomial")
```

## Example tree and polynomial  

Consider a three tip tree: 

```{r threeTipTree}
library(ape)
library(treenomial)

threeTipTree <- rtree(3, rooted = T)
plot.phylo(threeTipTree, use.edge.length = F, show.tip.label = F, direction = "downwards")

```

It's polynomial is x^3+xy+y which can equivalently be described with a coefficient matrix where the element in the ith row, jth column represents the y^(i-1) * x^(j-1) coefficient: 

```{r coeffMat}
treeToPoly(threeTipTree, varLabels = T)
```

Using the coefficients of the polynomials, distances between trees can be compared, below the two closest trees to a random target tree are found from a random sample:  

```{r plotExtremeTrees}
# random 12 tip target tree 
target <- rtree(12)

# random sample of 100 trees
sample <- rmtree(100,12)

minInfo <- plotExtremeTrees(target,sample, n = 2, comparison = "min", type = "d")
```



